### 2019-04-23 1-Preamble/4-Acknowledgement/acknowledgement.tex
Je suis tellement inefficace c'est triste

I would like to acknowledge and thank the following people, and also acknowledge the various funding who supported our works during the last three years.

@@

Then I would like to thank the dear members of my PhD jury.
I was lucky to be able to discuss with them on regular occasions, and these discussions have been very useful for some of our works.
Many thanks to \href{https://sites.google.com/site/vianneyperchet/}{Vianney Perchet} and \href{http://researchers.lille.inria.fr/~mitton/}{Nathalie Mitton} for accepting to review my thesis during the summer 2019, and for their very valuable comments.
Thanks to Patrick Maillé and Raphaël Féraud for following our work since Spring 2017,
and to Richard Combes as well for being part of my jury.
Some of their articles were strong inspirations for our own works, and I wish to also thank their co-authors, and in particular the Masters or PhD students, and useful discussions we had whenever we met, especially thanks to Claire Vernade, Réda Alami, Étienne Boursier, Frédéric Loge-Munerel and Pratik Gajane, and others I could not meet (\eg, Viktor Toldov or Robin Allesiardo).

Of course on a more personal level, my thoughts are strongly dedicated to my family.
%
I first want to thank my \href{https://paris-sorbonne.academia.edu/FBesson}{older brother Florian}, who has always been my main model and who has been of a constant support during all my studies. His whose solid and serious work kept pushing me further, some of our board games nights were brain teasers comparable with some maths problems solved for this thesis,
and I was happy to help him for the organization of the research colloque\footnote{Si vous l'avez manqué, un livre a été publié suite au colloque, \href{https://www.editions-vendemiaire.com/catalogue/a-paraitre/kaamelott-un-livre-d-histoire-florian-besson-et-justine-breton-dir/}{``Kaamelott, un livre d'histoire'', aux éditions Vandémiaire}.} on the \emph{Kaamelott} french TV show in February 2018, it was the funnier and most exciting research event I have been to in the last four years!
%
I also thank my younger brother Fabian, for being an inspiration for other (highly important) directions of my life.
Very warm thanks then go to Marine, to my grand-parents, Jeannine, Jean-Marc and Jacqueline, my uncle Jacques and aunt Bernadette, and my cousins.
Finally, I thank sincerely my parents, Christophe and Patricia, for being the best parents since the last 30 years, and for their vital moral support. \emph{Merci de nous avoir inculqué une si forte curiosité et d'être si gentils !}

@@

Thanks to my wonderful colleagues during these three years: Rémi$^*$, Pascal$^*$, Yves, Jacques, Ali, Haïfa, Marwa, Navik, Quentin$^*$, Vincent$^*$, Rami$^*$, Muhammad, Cristo$^*$, Adrien$^*$, Esteban$^*$, Ali, Éloïse$^*$, Nabil$^*$, Bastien$^*$, Corentin$^*$, Georgios, Morgane$^*$ in Rennes ; and Ronan, Odalric-Ambrym, Philippe, Olivier, Florian, Xuedong, Guillaume, Julien, Mathieu, Matteo, Mahsa, Nicolas in Lille.
I want to thank especially the ``coffee club'' members$^*$ in Rennes, for so many interesting discussions who where sometimes the highlights of our day-to-day life at the office.

@@

% I would like to conclude these acknowledgements by a special thought toward Lola, who was of great support on a daily basis, who marvelled me by her enthusiasm and dynamism, and who comforted me with her laughs and love, in the last year.
% \textcolor{white}{I love you Lola, would you like to live with me?}
%
% Je souhaite terminer ces remerciements par une pensée toute particulière pour Lola, qui m'a soutenu quotidiennement, qui m'a émerveillé par son enthousiasme et qui m'a réconforté par son amour, durant la dernière année.
% \textcolor{white}{Je t'aime Lola, veut-tu vivre avec moi ?}
%
\emph{Deseo terminar este agradecimiento con un pensamiento especial para Lola, quien me apoyó diariamente, me sorprendió con su entusiasmo y me consoló con sus risas y su amor, durante el año pasado.}
\textcolor{white}{Te quiero, Lola, ¿quieres vivir conmigo?}

>>>>

I would like to acknowledge and thank the following people, and also acknowledge the various funding who supported our works during the last three years and made this thesis possible.

@@

Then I would like to thank the dear members of my PhD jury.
I was lucky to be able to discuss with them on regular occasions, and these discussions have been very useful for some of our works.
Many thanks to \href{https://sites.google.com/site/vianneyperchet/}{Vianney Perchet} and \href{http://researchers.lille.inria.fr/~mitton/}{Nathalie Mitton} for accepting to review my thesis during the summer 2019, and for their very valuable comments.
Thanks to \href{http://perso.telecom-bretagne.eu/patrickmaille/}{Patrick Maillé} and \href{https://www.researchgate.net/profile/Raphael_Feraud/}{Raphaël Féraud} for following our work since Spring 2017,
and to \href{http://rcombes.supelec.free.fr/}{Richard Combes} as well for being part of my jury.
Some of their articles were strong inspirations for our own works, and I wish to also thank their co-authors, and in particular the Masters or PhD students, and useful discussions we had whenever we met, especially thanks to \href{https://www.cvernade.com/}{Claire Vernade}, \href{https://www.linkedin.com/in/reda-alami-067304109}{Réda Alami}, \href{https://www.researchgate.net/profile/Etienne_Boursier}{Étienne Boursier}, \href{https://www.researchgate.net/profile/Frederic_Loge_Munerel}{Frédéric Loge-Munerel} and \href{https://pratikgajane.wordpress.com/}{Pratik Gajane}, and others I could not meet (\eg, Viktor Toldov or \href{https://fr.linkedin.com/in/robin-allesiardo-phd-39221a92}{Robin Allesiardo}).

Of course on a more personal level, my thoughts are strongly dedicated to my family.
%
I first want to thank my \href{https://paris-sorbonne.academia.edu/FBesson}{older brother Florian}, who has always been my main model and who has been of a constant support during all my studies. His solid and serious work kept pushing me further, some of our board games nights were brain teasers comparable with some maths problems solved for this thesis,
and I was happy to help him for the organization of the \href{http://lettres.sorbonne-universite.fr/Kaamelott}{research colloque}\footnote{Un livre a aussi été publié suite \href{http://lettres.sorbonne-universite.fr/Kaamelott}{au colloque}, \href{https://www.editions-vendemiaire.com/catalogue/a-paraitre/kaamelott-un-livre-d-histoire-florian-besson-et-justine-breton-dir/}{``Kaamelott, un livre d'histoire'', aux éditions Vandémiaire}.} on the \emph{Kaamelott} french TV show in February 2017, it was \href{https://www.fabula.org/actualites/kaamelott-ou-la-relecture-de-l-histoire_75490.php}{the funniest} and \href{https://www.vousnousils.fr/2017/03/22/kaamelott-un-tremplin-vers-lhistoire-et-la-litterature-601428}{most exciting} \href{https://lpcm.hypotheses.org/10718}{research event} I have been to in the last four years!
%
I also thank my younger brother Fabian, for being an inspiration for other (highly important) directions of my life.
Very warm thanks then go to Marine, to my grand-parents, Jeannine, Jean-Marc and Jacqueline, my uncle Jacques and aunt Bernadette, and my cousins.
Finally, I thank sincerely my parents, Christophe and Patricia, for being the best parents since the last 30 years, and for their vital moral support. \emph{Merci de nous avoir inculqué une si forte curiosité et d'être si gentils !}

@@

Thanks to my wonderful colleagues during these three years: Rémi$^*$, Pascal$^*$, Yves, Jacques, Ali, Haïfa, Marwa, Navik, Quentin$^*$, Vincent$^*$, Rami$^*$, Muhammad, Cristo$^*$, Adrien$^*$, Esteban$^*$, Ali, Éloïse$^*$, Nabil$^*$, Bastien$^*$, Corentin$^*$, Georgios, Morgane$^*$ in Rennes ; and Ronan, Odalric-Ambrym, Philippe, Olivier, Florian, Xuedong, Guillaume, Julien, Mathieu, Matteo, Mahsa, Nicolas, Edouard in Lille.
I want to thank especially the ``coffee friends$^*$'' in Rennes, for the daily interesting discussions, which where sometimes the highlights of our life at the office.
%
Thanks to everybody who played \emph{futsal} on Wednesday lunch, I still have to progress immensely before I could pretend to be useful, but it's been a lot of fun to play with you guys!

@@

% I would like to conclude these acknowledgements by a special thought toward Lola, who was of great support on a daily basis, who marvelled me by her enthusiasm and dynamism, and who comforted me with her laughs and love, in the last year.
% \textcolor{white}{I love you Lola, would you like to live with me?}
%
% Je souhaite terminer ces remerciements par une pensée toute particulière pour Lola, qui m'a soutenu quotidiennement, qui m'a émerveillé par son enthousiasme et qui m'a réconforté par son amour, durant la dernière année.
% \textcolor{white}{Je t'aime Lola, veut-tu vivre avec moi ?}
%
\emph{Deseo terminar este agradecimiento con un pensamiento especial para Lola, quien me apoyó diariamente, me sorprendió con su entusiasmo y me consoló con sus risas y su amor, durante el año pasado.}\\
\indent
\textcolor{white}{$\heartsuit$ Te quiero, Lola, ¿quieres vivir conmigo?}

### 2019-04-23 2-Chapters/3-Chapter/chapter3.tex
Je suis tellement inefficace c'est triste

Any simulation library targeting multi-player bandit problems must implement at least another component:
a collision model, and a simulation loop that essentially looks like this.
\begin{itemize}
    \item First, initialize the MAB problem with $M$ players, and one or more cohorts of $M$ players (one player is one algorithm, usually $M$ times the same one),
    \item Then, for $t=1$ to $t=T$ (known before hand), repeat the following block (for each algorithm). Ask each player $\cA^{(j)}$ his chosen arm $A^{(j)}(t)$, then query the collision model\footnote{~The default collision model is the most widely studied in the literature, where a player encounters a collision (\ie, received a zero reward $Y^{(j)}(t)=0$) if she is not the only one to have chosen an arm $k=A^{(j)}(t)$, otherwise $Y^{(j)}(t)=r_{A^{(j)}(t)}(t)$ if she is the only one playing this arm.} to know which player will get a zero reward (for a collision) and which player will get a random reward from the environment. The
    sample (random) feedback $r_k(t)$ (\iid) from distributions $\nu_{k}$, and compute the rewards $Y^{(j)}(t)$ from the random feedback and the collision model. Finally feeds the observation $(A^{(j)}(t),Y^{(j)}(t))$ to the $M$ players,
    \item At the end, compute the accumulated reward, the regret, plot visualizations etc.
\end{itemize}

@@

\paragraph{What we did implement: \tettt{Joblib} for multi-core simulations.}
%
The first approach is to use multiple cores of the same machines, and because it is both the simplest and the less financially as well as ecologically costly, this is the approach implemented in SMPyBandits.
The machines I had access to during my thesis, either my own laptop or a workstation hosted the SCEE team in CentraleSupélec campus, were equipped with $i5$ or $i7$ Intel CPU with $4$ or $12$ cores.
%
As explained in the page \href{https://smpybandits.github.io/How_to_run_the_code.html}{\texttt{SMPyBandits.GitHub.io/How\_to\_run\_the\_code.html}}, we implemented in SMPyBandits an easy way to run any simulations on $n$ cores of a machine, using the \texttt{Joblib} \cite{joblib} library.
It is implemented in a completely transparent way, and if someone uses the command-line variable to configure experiments, using one core or all the cores of the machine one changes \texttt{N\_JOBS=1} to \texttt{N\_JOBS=-1}, like in this example.

>>>>

Any simulation library targeting multi-player bandit problems must implement at least another component:
a collision model, and a simulation loop that essentially looks like this.
\begin{itemize}
    \item First, initialize the MAB problem with $M$ players, and one or more cohorts of $M$ players (one player is one algorithm, usually $M$ times the same one),
    \item Then, for $t=1$ to $t=T$ (known before hand), repeat the following block (for each algorithm). Ask each player $\cA^{(j)}$ his chosen arm $A^{(j)}(t)$, then query the collision model\footnote{~The default collision model is the most widely studied in the literature, where a player encounters a collision (\ie, received a zero reward $Y^{(j)}(t)=0$) if she is not the only one to have chosen an arm $k=A^{(j)}(t)$, otherwise $Y^{(j)}(t)=r_{A^{(j)}(t)}(t)$ if she is the only one playing this arm.} to know which player will get a zero reward (for a collision) and which player will get a random reward from the environment. The
    sample (random) feedback $Y_k(t)$ (\iid) from distributions $\nu_{k}$, and compute the rewards $r^{(j)}(t)$ from the random feedback and the collision model. Finally feeds the observation $(A^{(j)}(t), Y_{A^{(j)}(t)}(t), r^{(j)}(t))$ to player $j$, for all the $M$ players,
    \item At the end, compute the accumulated reward, the regret, plot visualizations etc.
\end{itemize}

@@

\paragraph{What we did implement: \texttt{Joblib} for multi-core simulations.}
%
The first approach is to use multiple cores of the same machines, and because it is both the simplest and the less financially as well as ecologically costly, this is the approach implemented in SMPyBandits.
The machines I had access to during my thesis, either my own laptop or a workstation hosted the SCEE team in CentraleSupélec campus, were equipped with $i5$ or $i7$ Intel CPU with $4$ or $12$ cores.
%
As explained in the page \href{https://smpybandits.github.io/How_to_run_the_code.html}{\texttt{SMPyBandits.GitHub.io/How\_to\_run\_the\_code.html}}, we implemented in SMPyBandits an easy way to run any simulations on $n$ cores of a machine, using the \texttt{Joblib} \cite{joblib} library.
It is implemented in a completely transparent way, and if someone uses the command-line variable to configure experiments, using one core or all the cores of the machine one changes \texttt{N\_JOBS=1} to \texttt{N\_JOBS=-1}, like in this example.

### 2019-04-23 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Je suis tellement inefficace c'est triste

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $i$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

We further assume that there are $K \geq 2$ channels, $D \geq 0$ dynamic end-devices, and $S \geq 0$ static devices.
Furthermore, in channel $i \in \llbracket 1; K \rrbracket$ there are $0 \leq S_i \leq S$ static devices (so $S = \sum_{i=1}^{K} S_i$).

@@

\paragraph{Some assumptions on the network occupation}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall i, S_i > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

@@

We present three different policies that can be used to assess the efficiency of the learning algorithms presented later on.
The first one is naive but can be used in practice, while the two others are very efficient but require full knowledge on the system (\ie, an oracle) and are thus unpractical.
%
They are however useful for our numerical simulations, and will be used as references, in order to show that our MAB-based approaches quickly learn to perform almost optimally.

@@

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $i$, at that time slot.
The probability of successful transmission is computed as follows, because the $S_i$ static devices in each channel $i$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$:
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{i=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;i)}_{\text{No one else sent in channel}\; i} \; \underbrace{\Pr(\text{sent in channel}\,i)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $i$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_i}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{i=1}^{K} (1-p)^{S_i} .
\end{align}
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of $(S_i)_i$ the repartition of static devices, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle chooses a fixed configuration of dynamic devices, with $D_i$ devices affected to channel $i$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{i=1}^{K} \Pr(\text{success}|\text{sent in channel}\;i) \; \Pr(\text{sent in channel}\;i) \nonumber \\
    & = \sum_{i=1}^{K} \underbrace{(1 - p)^{D_i - 1}}_{\;\;D_i - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_i}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_i / D }_{\;\;\text{Sent in channel}\; i\;\;}.
\end{align}

Consequently, an optimal allocation vector $(D_1,\dots,D_{K})$ is a solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{i=1}^{K} D_i = D, \label{eq:41:eqCstr}\\
    & D_i \geq 0 \qquad \forall i\in\llbracket 1;K\rrbracket . \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

    It gives a closed form expression for the (unique) optimal solution $D_i^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.

    \begin{equation}\label{eq:41:Dilambda}
        D_i^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda e}{(1-p)^{S_i-1}} \right)-1 \right]\right)^{+} .
    \end{equation}
\end{proposition}

\begin{proof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_i\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall i\in\llbracket 1;K \rrbracket$. For such values for $D_i$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{i=1}^{K} D_i (1 - p)^{S_i + D_i -1}$ is concave as the sum of concave functions
    \footnote{~It is worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    \item
    More details are given in Section~\ref{sec:4:proofLagrangian} in the Appendix of this Chapter.
    % \hfill{}$\square$
    \end{itemize}
\end{proof}

Where in the equation~\eqref{eq:41:Dilambda}, $(a)^{+} = \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto x e^x$ on $\mathbb{R^+} = [0, +\infty)$ (it can be computed numerically in an efficient manner) \cite{Corless96}.
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint
\begin{equation}\label{eq:41:constraintLambda}
 \sum_{i=1}^{K} D_i^*(\lambda) = D.
\end{equation}

Equation \eqref{eq:41:constraintLambda} can be solved numerically, with simple one-dimensional root finding algorithms.
Solving the optimization problem provides the optimal real number value for $D_i^*$, which has to be rounded\footnote{~Any rounding choice will give about the same repartition, up-to a difference of only one device by channel, and so we chose to round from below for the first channels.} to find the optimal number of devices for channel $i$:
%
$\widehat{D_i} = \lfloor D_i^* \rfloor$ for $1 \leq i < K$, and $\widehat{D_{K}} = D - \sum_{i=1}^{K - 1} \widehat{D_i}$.

@@

We propose a \emph{sequential} approximation of the optimal policy:
the third solution is a sub-optimal naive policy, simple to set up, but also unpractical as it also needs an oracle.
End-devices are iteratively inserted in the channels with the lowest load (\ie, the index $i$ minimizing $S_i + D_i(\tau)$ at global time step $\tau$). Once the number of devices in each channel is computed, the probability of sending successfully a message is also given by equation \eqref{eq:41:prob_col}.
This is the policy that would be used by dynamic devices if they were inserted one after the other, and if they had a perfect knowledge of the channel loads.

@@

We include another simulation in Figure~\ref{fig:41:figure4appendix}, with a uniform repartition of static devices (\ie, $\forall i, S_i = S/K$), to check that learning approach (here, only \UCB)
also gives interesting gain of performance, and achieve a close-to optimal successful transmission rate.
Except for the limit case of $100\%$ of dynamic devices, which corresponds to Figures~\ref{fig:41:100intelligent} and \ref{fig:41:ch2_100}, where the uniform access performs as well as the optimal oracle solution, the MAB-based approach almost instantly outperforms the baseline.

>>>>

\paragraph{One gateway and many devices}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

We further assume that there are $K \geq 2$ channels, $D \geq 0$ dynamic end-devices, and $S \geq 0$ static devices.
Furthermore, in channel $k \in \llbracket 1; K \rrbracket$ there are $0 \leq S_k \leq S$ static devices (so $S = \sum_{k=1}^{K} S_k$).

@@

\paragraph{Some assumptions on the network occupation}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for IoT networks.
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to have an average of active devices not larger than the number of channels.

@@

We present three different policies that can be used to assess the efficiency of the learning algorithms presented later on.
The first one is naive but can be used in practice, while the two others are very efficient but require full knowledge on the system (\ie, an oracle) and are thus unpractical.
%
They are however useful for our numerical simulations, and are used as references, to show that our MAB-based approaches quickly learn to perform almost optimally.

@@

In this case, for one dynamic device, a successful transmission happens if it is the only device to choose channel $k$, at that time slot.
The probability of successful transmission is computed as follows, because the $S_k$ static devices in each channel $k$ are assumed to be independent, and static and dynamic devices are assumed to \emph{not} transmit at each time $t$ with a fixed probability $1-p$:
\begin{equation}
    \Pr(\text{success}|\text{sent}) = \sum_{k=1}^{K} \underbrace{\Pr(\text{success}|\text{sent in channel}\;k)}_{\text{No one else sent in channel}\; k} \; \underbrace{\Pr(\text{sent in channel}\,k)}_{= 1/K, \text{ by uniform choice}}
\end{equation}

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device}\;k\;\text{did not sent in}\;i) = \prod_{k=1}^{D-1} \Pr(\text{device}\;k\;\text{did not sent in}\;i)$. And $\Pr(\text{device}\;k\;\text{sent in}\;i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
This expression \eqref{eq:41:strategynaive} is constant (in time), and easy to compute numerically, but comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it
(maybe after a long enough initial learning period).

@@

We investigate here the optimal policy that can be achieved if the dynamic devices have a perfect knowledge of $(S_k)_k$ the repartition of static devices, and a fully centralized decision making\footnote{~This optimal policy needs an \emph{oracle} seeing the entire system, and affecting all the dynamic devices, once and for all, for a fixed stationary scenario, in order to avoid any signaling overhead. This is not possible in IoT contexts as several completely independent networks may operate in a single place.} is possible.
We want to find the stationary repartition of devices into channels that maximizes the probability of having a successful transmission.

If the oracle chooses a fixed configuration of dynamic devices, with $D_k$ devices affected to channel $k$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \Pr(\text{success}|\text{sent in channel}\;k) \; \Pr(\text{sent in channel}\;k) \nonumber \\
    & = \sum_{k=1}^{K} \underbrace{(1 - p)^{D_k - 1}}_{\;\;D_k - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_k}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_k / D }_{\;\;\text{Sent in channel}\; k\;\;}.
\end{align}

Consequently, an optimal allocation vector $(D_1,\dots,D_{K})$ is a solution of the following real-valued constraint optimization problem:
%
\begin{subequations}\label{eq:41:prob}
\begin{align}
    % \begin{cases}
    \underset{D_1,\dots,D_{K}}{\arg\max}\; & \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}, \label{eq:41:optPb}\\
    \text{such that}\;\; & \sum_{k=1}^{K} D_k = D, \label{eq:41:eqCstr}\\
    & D_k \geq 0 \qquad \forall k\in\llbracket 1;K\rrbracket . \label{eq:41:ineqCstr}
    % \end{cases}
\end{align}
\end{subequations}

@@

    It gives a closed form expression for the (unique) optimal solution $D_k^*(\lambda)$, depending on the system parameters, and the unknown Lagrange multiplier $\lambda \in \mathbb{R}$.

    \begin{equation}\label{eq:41:Dilambda}
        D_k^*(\lambda) = \left(\frac{1}{\log(1-p)}\left[ \mathcal{W}\left(\frac{\lambda e}{(1-p)^{S_k-1}} \right)-1 \right]\right)^{+} .
    \end{equation}
\end{proposition}

\begin{proof}
\begin{itemize}
    \item
    In a realistic scenario, we can assume that $D_k\leq \frac{-2}{\ln\left(1-p\right)} \approx \frac{2}{p},\quad \forall k\in\llbracket 1;K \rrbracket$. For such values for $D_k$, the objective function $f: (D_1, \dots, D_{K}) \mapsto \sum_{k=1}^{K} D_k (1 - p)^{S_k + D_k -1}$ is concave as the sum of concave functions
    \footnote{~It is worth noting that $f$ is neither concave nor quasi-concave on $[0,\infty)^{K}$ \cite{Luenberger68,Yaari77}.}.
    \item
    The Lagrange multipliers method can be applied to the optimization problem \eqref{eq:41:optPb}, with a concave objective function $f$, linear equality constraints \eqref{eq:41:eqCstr} and linear inequality constraints \eqref{eq:41:ineqCstr}. The strong duality condition is satisfied in this case \cite{BoydVanderberghe04}, so finding the saddle points will be enough to find the maximizers.
    \item
    More details are given in Section~\ref{sec:4:proofLagrangian} in the Appendix of this Chapter.
    % \hfill{}$\square$
    \end{itemize}
\end{proof}

Where in the equation~\eqref{eq:41:Dilambda}, $(a)^{+} = \max(a,0)$, and $\mathcal{W}$ denotes the $\mathcal{W}$-Lambert function which is the reciprocal bijection of $x \mapsto x e^x$ on $\mathbb{R^+} = [0, +\infty)$ (it can be computed numerically in an efficient manner) \cite{Corless96}.
Moreover, condition \eqref{eq:41:eqCstr} implies that the Lagrange multiplier $\lambda$ is the solution of the constraint
\begin{equation}\label{eq:41:constraintLambda}
    \sum_{k=1}^{K} D_k^*(\lambda) = D.
\end{equation}

Equation \eqref{eq:41:constraintLambda} can be solved numerically, with simple one-dimensional root finding algorithms.
Solving the optimization problem provides the optimal real number value for $D_k^*$, which has to be rounded\footnote{~Any rounding choice will give about the same repartition, up-to a difference of only one device by channel, and so we chose to round from below for the first channels.} to find the optimal number of devices for channel $k$:
%
$\widehat{D_k} = \lfloor D_k^* \rfloor$ for $1 \leq k < K$, and $\widehat{D_{K}} = D - \sum_{k=1}^{K - 1} \widehat{D_k}$.

@@

We propose a \emph{sequential} approximation of the optimal policy:
the third solution is a sub-optimal naive policy, simple to set up, but also unpractical as it also needs an oracle.
End-devices are iteratively inserted in the channels with the lowest load (\ie, the index $k$ minimizing $S_k + D_k(\tau)$ at global time step $\tau$). Once the number of devices in each channel is computed, the probability of sending successfully a message is also given by equation \eqref{eq:41:prob_col}.
This is the policy that would be used by dynamic devices if they were inserted one after the other, and if they had a perfect knowledge of the channel loads.

@@

We include another simulation in Figure~\ref{fig:41:figure4appendix}, with a uniform repartition of static devices (\ie, $\forall k, S_k = S/K$), to check that learning approach (here, only \UCB)
also gives interesting gain of performance, and achieve a close-to optimal successful transmission rate.
Except for the limit case of $100\%$ of dynamic devices, which corresponds to Figures~\ref{fig:41:100intelligent} and \ref{fig:41:ch2_100}, where the uniform access performs as well as the optimal oracle solution, the MAB-based approach almost instantly outperforms the baseline.
